State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-07-12 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
Revise document to smooth out changes made by JC.  Ready for double blind next.

% \author[1,*]{W.\ Spencer Smith}
% \author[1]{Ao Dong}
% \author[1]{Jacques Carette}
% \author[2]{Michael D.\ Noseworthy}

% \affil[1]{McMaster University, Computing and Software Department, {Canada}}
% \affil[2]{McMaster University, Electrical \& Computer Engineering Department,
% {Canada}}
% \affil[*]{Corresponding Author}

\abstract{
We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/ verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each. Using the Analytic Hierarchy Process (AHP) on the
quantitive data, we ranked the MI software.  The top five are \textit{3D
Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
\textit{ParaView}.  This is consistent with the community's, with four of
these also appearing in the top five using GitHub metrics (stars-per-year).  
In general, the quality and quantity of documentation present in the
project correlates quite well with its popularity.
%We observed 88\% of the documentation artifacts
%recommended by research software development guidelines.  However, the current
%state of the practice deviates from the existing guidelines because of the
%rarity of some recommended artifacts (like test plans, requirements
%specification, code of conduct, code style guidelines, product roadmaps, and
%Application Program Interface (API) documentation).
}

@@

% RQ1 to RQ4
% Some of Section 1 (Introduction)
% Sections 2 to 6 (includes looking at artifacts observed in repos)
% Some of Sections 11 to 13 (Threats to Validity, Future Work, Conclusions)

@@

We study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide information for diagnostic, analytic, and medical
applications. Given its importance, we want to understand the merits and
drawbacks of the current development processes, tools, and methodologies. We
use a software engineering lens to assess the quality of existing MI software
and provide recommendations for future development.

@@

%We have a standard set of questions designed to assess the qualities of any
%research software project \cite{SmithEtAl2021, SmithAndMichalski2022}. We
%particularly focus on \emph{software artifactcs}, i.e.
%the documents, scripts and code that constitutes a software development
%project. Example artifacts include requirements, specifications, user manuals,
%unit tests, system tests, usability tests, build scripts, API (Application
%Programming Interface) documentation, READMEs, license documents, process
%documents, and code.  %Once we have learned what MI developers do, we then put
%% this information in context by contrasting MI software against the trends shown
%% by developers in other research software communities.

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
\textbf{open source} software projects exist? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Establish
quantitative measurements of those project's development practices. From
these measures, which of those projects follow best practices?
(Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\end{enumerate}

@@

We identify a list of potential packages (through online searches) which is
then filtered and vetted by a domain expert. We aim for roughly $30$
packages. For each remaining package, we measure its qualities by filling in a
grading template~\cite{SmithEtAl2021}.  This data is used to rank the projects
with AHP.  Further details on interaction with the domain expert, software
qualities, grading the software and AHP are summarizes below and in longer form
in~\cite{SmithEtAl2024_MI_SOP}.

@@

The Domain Expert vets the proposed list. For too many domains, there are
online resources that are wildly inaccurate, which only an expert can correct.
(Details of our domain expert removed for double-blind).
%The Domain Experts vet the proposed list of software packages and the AHP
%ranking.  Doing these tasks with non-experts runs the risk of trusting
%inaccurate resources and overlooking the tacit knowledge of an expert. For the
%current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
%University, Co-Director of the McMaster School of Biomedical Engineering, and
%Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
%Hamilton, Ontario, Canada.

@@

We use an existing template~\cite{SmithEtAl2021} that is designed to measure
the aforementioned qualities. To stay within our given measurement time frame,
each package gets up to five hours of time.  Project developers can be
contacted for help regarding installation, if necessary, but we impose a cap of
about two hours on the installation process.
Figure~\ref{fg_grading_template_example} shows an excerpt of the measurement
spreadsheet.  It includes a column for each measured software package.
% \cite{Dong2021-Data} provides the full set of measurement data.
(The full data is available on Mendeley; link will be provided after
refereeing.)

@@

The full template consists of 108 questions over 9 qualities.  These
questions are designed to be unambiguous, quantifiable, and measurable
with constrained time and domain knowledge.
%We
%designed the questions to be unambiguous, quantifiable, and measurable with
%limited time and domain knowledge. We group the measures under headings for each
%quality, and one for summary information. The summary information (shown in
%Figure~\ref{fg_grading_template_example}) is the first section of the template.
%This section summarizes general information, such as the software name, purpose,
%platform, programming language, publications about the software, the first
%release and the most recent change date, website, source code repository of the
%product, number of developers, etc.  We only measure open source software that
%is ready for public use by other researchers (following the definition of the
%public category from \cite{GewaltigAndCannon2012}). Information in the summary
%section sets the context for the project, but it does not directly affect the
%grading scores.

%For measuring each quality, we ask several questions and the typical answers are
%among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
%string, a date, a set of strings, etc. 
The grader, after answering questions for each quality assigned an overall
score (between 1 and 10) based on the answers.  Several of the qualities use
the word ``surface'' to highlight that these particular qualities are a shallow
measure.  For example, we do not do user studies for measuring usability.
Instead, we are look signs that the developers considered usability.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
details on quality measurement are provided in~% Smith et al.\ (2024)
\cite{SmithEtAl2024_MI_SOP}.

@@

The projects are sorted in descending order of lines of code.
We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. The date of the last
update is the date of the latest update, at the time of measurement. We found
funding information (Fnd) for only eight projects.  For the Number Of
Contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available on \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1}
{Mendeley Data}.

@@

The scores are based on the ease of following the installation instructions,
and automated installation and uninstallation process. There were no issues
for all but the bottom four, just various degrees of ease and automation.

\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
installation problems. We were not able to install them, even after a reasonable
amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
from the source code, but we were able to proceed with measuring other qualities
using a deployed on-line version for \textit{dwv}, and a VM version for
\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
dependency, and thus could not measure reliability nor robustness for it.

@@

The packages with higher scores for correctness and verifiability (see
Figure~\ref{fg_correctness_verifiability_scores}) used a wider array of
techniques to improve correctness, and had better documentation to witness
this. For instance, we looked for evidence of unit testing, and only evidence
for about half of the projects. We identified five projects using CI/CD
tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}.

@@

Even for some projects with well-organized documentation, requirements
specifications and theory manuals were still missing.
The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

@@

Figure~\ref{fg_reliability_scores} shows the our results.  We were able to
follow the steps in the tutorials that existed (seven packages had them.)
However
\textit{GATE} could not open macro files and became unresponsive several times,
without any descriptive error message. We found that \textit{Drishti} crashed
when loading damaged image files, without showing any descriptive error message.

@@

Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
The packages with higher scores elegantly handled unexpected or unanticipated
inputs, typically showing a clear error message. We may have underscored
\textit{OHIF Viewer}, since we needed further customization to load
data.

@@

Figure~\ref{fg_maintainability_scores} shows the ranking results for
maintainability. We gave \textit{3D Slicer} the highest score because we found
it to have the most comprehensive artifacts. Only a few of the 29 projects had a
product, developer's manual, or API documentation, and only \textit{3D Slicer},
\textit{ImageJ}, \textit{Fiji} included all three documents 
(see Table~\ref{tab_maintainability_docs} the full data).  Moreover,
\textit{3D Slicer} has a much higher percentage of closed issues (92\%) compared
to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).

@@

We first compare our ranking to a (proxy for) the community's ranking. We then
compare the state of the practice for MI with that of other research software.
In particular we provide details on recommended artifacts but rarely observed
for MI software. Section~\ref{sec_threats_to_validity} presents threats to the
validity of our data and conclusions.

@@

We ranked some popular packages fairly low, such as \textit{dwv}. This is
because we were unable to build it locally, even though we followed its
installation instructions. But we were able to use its web
version for the rest of the measurements. Additionally, this version 
did not detect a broken DICOM file and instead displayed a blank image
(Section~\ref{sec_result_robustness}).
\textit{DICOM Viewer} ranked low as we were unable to install the
NextCloud platform.

@@

We use nine research software development guidelines to measure what
software artifacts are present in MI software, versus those recommended.
These guidelines are:
\begin{itemize}
\item United States Geological Survey Software Planning Checklist
\cite{USGS2019},
\item DLR (German Aerospace Centre) Software Engineering Guidelines
\cite{TobiasEtAl2018}, 
\item Scottish Covid-19 Response Consortium Software Checklist
\cite{BrettEtAl2021},
\item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
Policies \cite{SmithAndRoscoe2018},
\item Trilinos Developers Guide \cite{HerouxEtAl2008},
\item EURISE (European Research Infrastructure Software Engineers') Network
Technical Reference \cite{ThielEtAl2020},
\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
Guidelines for Software Quality \cite{vanGompelEtAl2016}, and
\item A Set of Common Software Quality Assurance Baseline Criteria for Research
Projects \cite{OrvizEtAl2017}.
\end{itemize}

@@

Table~\ref{artifactspresent} presents our measurements for MI software.
It groups the artifacts by frequency
into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
show the details on which projects use which types of artifacts for documents
related to maintainability and visibility, respectively.

Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does no imply that these
oft recommended artifacts are the most important. Guidelines are often brief,
to encourage adoption, and thus even guidelines that mention the need for
installation instructions rarely mention uninstallation instructions.
Two items in Table~\ref{artifactspresent} do not appear in any guidelines:
\emph{Troubleshooting guide} and \emph{Developer's manual}.  However the
information within these documents overlaps with
the recommended artifacts.  Troubleshooting information often can be found in
a User Manual, while the information in a ``Developer's Manual'' is often
scattered amongst many other documents.
Three of the 26 recommended artifacts were never observed:
i) Uninstall, ii) Test plans, and iii) Requirements. It is possible that
some of these were created but never put under version control.

It is unfortunately common for research software to neglect requirements
documentation, and MI software is no exception.
Although such documentation is recommended by some
\cite{TobiasEtAl2018, HerouxEtAl2008, SmithAndKoothoor2016}, in practice
this is rare~\cite{HeatonAndCarver2015}. Sanders and Kelly
\cite{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
found that none of the scientists created requirements specifications, unless
regulations in their field mandated such a document.  Requirements are the least
commonly produced type of documentation for research software in general
\cite{Nguyen-HoanEtAl2010}.

This is quite unfortunate as when surveyed, research software developers
identify the lack of requirements and requirements management as accounting
gets mentioned the most often and accounts for 23\% of the technical problems
\cite{WieseEtAl2019}.  There is unfortunately widespread perception that
up-front requirements are impossible for research software
\cite{CarverEtAl2007, SegalAndMorris2008}. Fortunately a more agile approach
to requirements is feasible~\cite{Smith2016}, and research-software specific
templates exist~\cite{SmithEtAl2007}.

@@

MI projects fall somewhat short of recommended best practices, but are not
alone amongst research software projects. This gap has been
documented before~\cite{Storer2017,JohansonAndHasselbring2018}, is known
to cause sustainability and reliability
problems \cite{FaulkEtAl2009}, and to waste development effort%
~\cite{deSouzaEtAl2019}.

@@

A study is reliable if repeated by another researcher, using the same
methodology, would lead to the same results~\cite{RunesonAndHost2009}.
We identify the following threats:

@@

Construct validity is when the adopted metrics represent what they are intended
to measure~\cite{RunesonAndHost2009}. We have identified the following
potential issues:
\begin{itemize}
\item We make indirect measurement of software qualities since meaningful direct
measures for qualities like maintainability, reusability and verifiability, are
unavailable.  We follow the usual assumption that developers achieve higher
quality by following procedures and adhering to standards \cite[p.\
112]{VanVliet2000}.
\item We could not
install or build \textit{dwv}, \textit{GATE}, and \textit{DICOM Viewer}. We used
a deployed on-line version for \textit{dwv}, a VM version for \textit{GATE}, but
no alternative for \textit{DICOM Viewer}.
\item Robustness measurements involve only two pieces of data, leading
to limited variation in the robustness scores
(Figure~\ref{fg_robustness_scores}). Our measurement-time budget limited
what we could achieve her.
\item Our maintainability proxies (higher ratio of comments to source,
high percentage of closed issues) have not been validated.
\item While smaller modules tend to be easier to reuse, small modules are
not necessarily good modules, nor understandable modules.
\item The understandability measure relies on 10 random source code files, but
the 10 files will not necessarily be representative. 
\item Our overall AHP ranking makes the unrealistic assumption of equal
weighting.
\item We approximated popularity by stars and watches.
\item Table~\ref{Tbl_Guidelines} required judgement as
not all guidelines use the same names for artifacts that contain
essentially the same information.
\end{itemize}

@@

\begin{itemize}
\item Our search (Section~\ref{ch_results}) may
have missed a relevant package.
\item Our methodology assumes that development activities
will leave a trace in the repositories, but this is not necessarily true. For
instance, we saw little evidence of requirements
(Section~\ref{Sec_CompareArtifacts}), but teams might keep this kind of
information outside their repos.
\end{itemize}

@@

If the results of a study can be generalized to other situations, then the
study is externally valid \cite{RunesonAndHost2009}.  In other words, cannot
generalize our results if the development of MI software is fundamentally
different from other research software.  Although there are differences, like
the importance of data privacy for MI data, we found the approach to developing
LBM software \cite{SmithEtAl2024} and MI software to be similar.  Except for
the domain specific aspects, the trends observed in the current study are
similar to that for other research software.

@@

Our analysis of the state of the practice for the MI domain along nine
software qualities strongly indicates that ``higher quality'' is
consistent with community ranking proxies. It is important to note that
our quality measures are, on purpose, rather shallow. This undescores the
importance that \emph{first impressions} of any software has on its
adoption.

@@

While we did find a reasonable amount of documentation, especially when
consider all MI projects, there were definitely some holes. Some important
documentation (test plans and requirements documentation) was missing,
and other (contributors' guide, code of conduct, code style guidelines, product
roadmap, design documentation, and API documentation
(Section~\ref{Sec_CompareArtifacts}) were rare.

>>>>

\abstract{
We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/ verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each. Using the Analytic Hierarchy Process (AHP) on the
quantitative data, we ranked the MI software.  The top five are \textit{3D
Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
\textit{ParaView}.  This is consistent with the community's view, with four of
these also appearing in the top five using GitHub metrics (stars-per-year).  
In general, the quality and quantity of documentation present in a
project correlates quite well with its popularity.
}

@@

@@

We study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide critical information for diagnostic, analytic, and medical
applications. Given its importance, we want to understand the merits and
drawbacks of the current development processes, tools, and methodologies. We
use a software engineering lens to assess the quality of existing MI software.

@@

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
\textbf{open source} software projects exist? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Based on
quantitative measurements of each project's development practices, which
projects follow best practices? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts (documents, scripts and code) present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\end{enumerate}

@@

We identify a list of potential packages (through online searches) which is then
filtered and vetted by a domain expert. We aim for roughly $30$ packages. For
each remaining package, we measure its qualities by filling in a grading
template~\cite{SmithEtAl2021}.  This data is used to rank the projects with the
Analytic Hierarchy Process (AHP).  We summarize further details on the
interaction with the domain expert, software qualities, grading the software and
AHP below and in longer form in Smith et al.\
(2024)~\cite{SmithEtAl2024_MI_SOP}.

@@

The Domain Expert vets the proposed list because online resources can be
inaccurate.  The expert also vets the AHP ranking.  For the current assessment,
our Domain Expert is [Details of our domain expert removed for double-blind].
%%DOUBLE BLIND %our Domain Expert (and paper co-author) is Dr.\ Michael
%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
%University, Co-Director of the McMaster School of Biomedical Engineering, and
%Director of Medical Imaging Physics and Engineering at St.\ Joseph's
%Healthcare, Hamilton, Ontario, Canada.

@@

We use an existing template~\cite{SmithEtAl2021} that is designed to measure the
aforementioned qualities. To stay within our given measurement time frame, each
package gets up to five hours of time.  Project developers can be contacted for
help regarding installation, if necessary, but we impose a cap of about two
hours on the installation process.  Figure~\ref{fg_grading_template_example}
shows an excerpt of the measurement spreadsheet.  The rows are the measures and
the columns correspond to the software packages.  [The full data is available on
Mendeley; link will be provided after refereeing.] 
% \cite{Dong2021-Data}
%provides the full set of measurement data.   %%DOUBLE BLIND

@@

The full template consists of 108 questions over 9 qualities.  These
questions are designed to be unambiguous, quantifiable, and measurable
with constrained time and domain knowledge.

The grader, after answering questions for each quality assigns an overall
score (between 1 and 10) based on the answers.  Several of the qualities use
the word ``surface'' to highlight that these particular qualities are a shallow
measure.  For example, usability is not measured using user studies.
Instead, we look for signs that the developers considered usability.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
details on quality measurement are provided in Smith et al.\ (2024)
\cite{SmithEtAl2024_MI_SOP}.

@@

The projects are sorted in descending order of lines of code.  We found the
initial release dates (Rlsd) for most projects and marked the two unknown dates
with ``?''. The date of the last update is the date of the latest update, at the
time of measurement. We found funding information (Fnd) for only eight projects.
For the Number Of Contributors (NOC) we considered anyone who made at least one
accepted commit as a contributor. The NOC is not usually the same as the number
of long-term project members, since many projects received change requests and
code from the community.  With respect to the OS, 25 packages work on all three
OSs: Windows (W), macOS (M), and Linux (L). Although the usual approach to
cross-platform compatibility was to work natively on multiple OSes, five
projects achieved platform-independence via web applications. The full
measurement data for all packages is available on [removed for blind review]
%\href{https://data.mendeley.com/datasets/k3pcdvdzj2/1} {Mendeley Data}.  
%% DOUBLE BLIND

@@

The scores are based on the ease of following the installation instructions,
and automated installation and uninstallation process. There were no issues
for all but the bottom four, just various degrees of ease and automation.
\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
installation problems. We were not able to install them, even after a reasonable
amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
from the source code, but we were able to proceed with measuring other qualities
using a deployed on-line version for \textit{dwv}, and a virtual machine version for
\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
dependency, and thus could not measure reliability nor robustness for it.

@@

The packages with higher scores for correctness and verifiability (see
Figure~\ref{fg_correctness_verifiability_scores}) used a wider array of
techniques to improve correctness, and had better documentation to witness this.
For instance, we looked for evidence of unit testing, and found evidence for
only about half of the projects. We identified five projects using continuous
integration tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji},
\textit{dwv}, and \textit{OHIF Viewer}.

@@

Even for projects with well-organized documentation, requirements
specifications and theory manuals were still missing.
The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

@@

Figure~\ref{fg_reliability_scores} shows our results.  We were able to follow
the steps in the tutorials that existed (seven packages had them.) However,
\textit{GATE} could not open macro files and became unresponsive several times,
without any descriptive error message. We found that \textit{Drishti} crashed
when loading damaged image files, without showing any descriptive error message.

@@

Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
The packages with higher scores gracefully handled unexpected or unanticipated
inputs, typically showing a clear error message. We may have underscored
\textit{OHIF Viewer}, since we needed further customization to load
data.

@@

Figure~\ref{fg_maintainability_scores} shows the ranking results for
maintainability. We gave \textit{3D Slicer} the highest score because we found
it had the most comprehensive artifacts. Only a few of the 29 projects had a
product, developer's manual, or API (Application Programming Interface)
documentation, and only \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}
included all three documents (see Table~\ref{tab_maintainability_docs} for the
full data).  Moreover, \textit{3D Slicer} has a much higher percentage of closed
issues (92\%) compared to \textit{ImageJ} (52\%) and \textit{Fiji} (64\%).

@@

We first compare our ranking to a (proxy for) the community's ranking. We then
compare the state of the practice for MI with that of other research software.
In particular we provide details on recommended artifacts that are rarely observed
for MI software. Section~\ref{sec_threats_to_validity} presents threats to the
validity of our data and conclusions.

@@

We ranked some popular packages fairly low, such as \textit{dwv}. This is
because we were unable to build it locally, even though we followed its
installation instructions. However, we were able to use its web
version for the rest of the measurements. Additionally, this version 
did not detect a broken DICOM file and instead displayed a blank image
(Section~\ref{sec_result_robustness}).
\textit{DICOM Viewer} ranked low as we were unable to install the
NextCloud platform.

@@

We use nine research software development guidelines to compare recommended 
software artifacts versus those present in MI software.
These guidelines are:
\begin{itemize}
\item United States Geological Survey Software Planning Checklist
\cite{USGS2019},
\item DLR (German Aerospace Centre) Software Engineering Guidelines
\cite{TobiasEtAl2018}, 
\item Scottish Covid-19 Response Consortium Software Checklist
\cite{BrettEtAl2021},
\item Good Enough Practices in Scientific Computing \cite{WilsonEtAl2016},
\item xSDK (Extreme-scale Scientific Software Development Kit) Community Package
Policies \cite{SmithAndRoscoe2018},
\item Trilinos Developers Guide \cite{HerouxEtAl2008},
\item EURISE (European Research Infrastructure Software Engineers') Network
Technical Reference \cite{ThielEtAl2020},
\item CLARIAH (Common Lab Research Infrastructure for the Arts and Humanities)
Guidelines for Software Quality \cite{vanGompelEtAl2016}, and
\item A Set of Common Software Quality Assurance Baseline Criteria for Research
Projects \cite{OrvizEtAl2017}.
\end{itemize}

@@

Table~\ref{artifactspresent} presents our measurements for MI software.
The table groups the artifacts by frequency
into categories of common (20 to 29 ($>$67\%) packages), uncommon (10 to 19
(33-67\%) packages), and rare (1 to 9 ($<$33\%) packages).
Tables~\ref{tab_maintainability_docs} and~\ref{tab_Visibility/Transparency_docs}
show the details on which projects use which types of artifacts for documents
related to maintainability and visibility, respectively.

Note that ``popularity'' in Table~\ref{Tbl_Guidelines} does no imply that these
oft recommended artifacts are the most important. Guidelines are often brief,
to encourage adoption, and thus even guidelines that mention the need for
installation instructions rarely mention uninstallation instructions.
Two items in Table~\ref{artifactspresent} do not appear in any guidelines:
\emph{Troubleshooting guide} and \emph{Developer's manual}.  However the
information within these documents overlaps with
the recommended artifacts.  Troubleshooting information often can be found in
a User Manual, while the information in a ``Developer's Manual'' is often
scattered amongst many other documents.

Three of the 26 recommended artifacts were never observed in the MI software:
i) Uninstall, ii) Test plans, and iii) Requirements. It is possible that
some of these were created but never put under version control.

Neglecting requirements documentation is unfortunately common for research
software, and MI software is no exception to this trend.  Although such
documentation is recommended by some \cite{TobiasEtAl2018, HerouxEtAl2008,
SmithAndKoothoor2016}, in practice this is rare~\cite{HeatonAndCarver2015}.
Sanders and Kelly \cite{SandersAndKelly2008} interviewed 16 scientists from 10
disciplines and found that none of the scientists created requirements
specifications, unless regulations in their field mandated such a document.
Requirements are the least commonly produced type of documentation for research
software in general \cite{Nguyen-HoanEtAl2010}.

This is unfortunate as when scientific developers are surveyed on their pain
points, Wiese et al.~\cite{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts them,
accounting for 23\% of the technical problems reported by study participants.
Further adding to the misfortune, there is a widespread perception that up-front
requirements are impossible for research software \cite{CarverEtAl2007,
SegalAndMorris2008}. Fortunately a more agile approach to requirements is
feasible~\cite{Smith2016}, and research-software specific templates
exist~\cite{SmithEtAl2007}.

@@

MI projects fall somewhat short of recommended best practices, but are not alone
amongst research software projects. This gap has been documented
before~\cite{Storer2017,JohansonAndHasselbring2018}, and is known to cause
sustainability and reliability problems \cite{FaulkEtAl2009}, and to waste
development effort~\cite{deSouzaEtAl2019}.

@@

A study is reliable if repeating it by another researcher, using the same
methodology, would lead to the same results~\cite{RunesonAndHost2009}.
We identify the following threats:

@@

Construct validity is when the adopted metrics represent what they are intended
to measure~\cite{RunesonAndHost2009}. We have identified the following
potential issues:
\begin{itemize}
\item We make indirect measurement of software qualities since meaningful direct
measures for qualities like maintainability, reusability and verifiability, are
unavailable.  We follow the usual assumption that developers achieve higher
quality by following procedures and adhering to standards \cite[p.\
112]{VanVliet2000}.
\item We could not install or build \textit{dwv}, \textit{GATE}, and
\textit{DICOM Viewer}. We used a deployed on-line version for \textit{dwv}, a
virtual machine version for \textit{GATE}, but no alternative for \textit{DICOM
Viewer}.
\item Robustness measurements involve only two pieces of data, leading
to limited variation in the robustness scores
(Figure~\ref{fg_robustness_scores}). Our measurement-time budget limited
what we could achieve here.
\item Our maintainability proxies (higher ratio of comments to source,
high percentage of closed issues) have not been validated.
\item While smaller modules tend to be easier to reuse, small modules are
not necessarily good modules, nor understandable modules.
\item The understandability measure relies on 10 random source code files, but
the 10 files will not necessarily be representative. 
\item Our overall AHP ranking makes the unrealistic assumption of equal
weighting.
\item We approximated popularity by stars and watches.
\item Table~\ref{Tbl_Guidelines} required judgement as
not all guidelines use the same names for artifacts that contain
essentially the same information.
\end{itemize}

@@

\begin{itemize}
\item Our search (Section~\ref{ch_results}) could
have missed a relevant package.
\item Our methodology assumes that development activities
will leave a trace in the repositories, but this is not necessarily true. For
instance, we saw little evidence of requirements
(Section~\ref{Sec_CompareArtifacts}), but teams might keep this kind of
information outside their repos.
\end{itemize}

@@

If the results of a study can be generalized to other situations, then the
study is externally valid \cite{RunesonAndHost2009}.  In other words, we cannot
generalize our results if the development of MI software is fundamentally
different from other research software.  Although there are differences, like
the importance of data privacy for MI data, we found the approach to developing
LBM software \cite{SmithEtAl2024} and MI software to be similar.  Except for
the domain specific aspects, the trends observed in the current study are
similar to that for other research software.

@@

Our analysis of the state of the practice for MI domain along nine software
qualities strongly indicates that ``higher quality'' is consistent with
community ranking proxies. Although our quality measures are rather shallow, we
see this as an advantage. The shallow measures are a proxy for the importance of
\emph{first impressions} for software adoption.

@@

While we did find a reasonable amount of documentation, especially when
consider all MI projects, there were definitely some holes. Some important
documentation (test plans and requirements documentation) was missing,
and other (contributors' guide, code of conduct, code style guidelines, product
roadmap, design documentation, and API documentation)
(Section~\ref{Sec_CompareArtifacts}) were rare.
